---
title: "1.3 Assigning TF to cell barcodes for scTF-seq data"
author: "Vincent Gardeux"
date: "2023/11/23"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "./")
getwd()
```

## Libraries & functions

First, I'm loading the required libraries & functions

```{r}
suppressPackageStartupMessages(library(Seurat)) # For handling 10x h5 file
suppressPackageStartupMessages(library(data.table)) # For fread/fwrite functions
suppressPackageStartupMessages(library(SamSPECTRAL)) # For the kneepoint algorithm
suppressPackageStartupMessages(library(Rfast)) # For colnth function
suppressPackageStartupMessages(library(crayon)) # Just for bolding the console output :D

cat(bold("Seurat"), "version", as.character(packageVersion("Seurat")), "\n")
cat(bold("data.table"), "version", as.character(packageVersion("data.table")), "\n")
cat(bold("SamSPECTRAL"), "version", as.character(packageVersion("SamSPECTRAL")), "\n")
cat(bold("Rfast"), "version", as.character(packageVersion("Rfast")), "\n")

## Kneepoint filtering and TF assignment
filtering_kneepoint <- function(exp_cellranger, exp_tf_cell, exp_out_seurat){
  ## Cell Ranger
  data.cell_ranger <- Read10X_h5(filename = exp_cellranger, use.names = F)
  
  ## Cell Ranger barcodes
  data.barcodes <- unlist(data.cell_ranger@Dimnames[2])
  data.barcodes <- gsub(pattern = "-1", replacement = "", x = data.barcodes)
  data.cell_ranger@Dimnames[2] <- list(data.barcodes)
  message(length(data.barcodes), " cell barcodes in cell ranger.")
  
  ## Create Seurat object
  data.seurat <- CreateSeuratObject(data.cell_ranger)
  rm(data.cell_ranger)
  
  # 10x
  data.tf_read <- fread(exp_tf_cell, data.table = F)
  rownames(data.tf_read) <- data.tf_read$TFName
  data.barcodes_tf <- colnames(data.tf_read)[3:ncol(data.tf_read)]
  message(length(data.barcodes_tf), " cell barcodes in TF/Cell")
  
  # Overlap
  common.barcodes <- intersect(data.barcodes_tf, data.barcodes)
  data.tf_read <- data.tf_read[,common.barcodes] # Filtered TF Matrix
  message(length(common.barcodes), " cell barcodes are overlapping.")

  # Calculate rate of main TF
  nCounts <- colSums(data.tf_read)
  nMax <- apply(data.tf_read, 2, max) 
  TFs.MaxRate <- nMax / nCounts
  TFsBC.MaxRate.cutoff <- TFs.MaxRate[nCounts > 5] # Select cells with at least 5 reads
  
  # Knee point cutoff detection
  TFsBC.MaxRate.cutoff_order <- TFsBC.MaxRate.cutoff[order(TFsBC.MaxRate.cutoff, decreasing = T)] # order from 1 to 0 
  cutoff <- kneepointDetection(TFsBC.MaxRate.cutoff_order)

  # Count how many cells are passing the filtering
  filtered_barcodes <- common.barcodes[TFs.MaxRate > TFsBC.MaxRate.cutoff_order[cutoff$MinIndex] & nCounts > 5]
  message(length(filtered_barcodes), " cells are passing the kneepoint filtering")
  
  # Subset cell_ranger matrix
  data.seurat <- data.seurat[,filtered_barcodes]
  data.tf_read <- data.tf_read[,filtered_barcodes]
  
  # Assign TFs
  TF <- c()
  nMax <- apply(data.tf_read, 2, max)
  for(i in 1:ncol(data.tf_read)){
    TF <- c(TF, rownames(data.tf_read)[data.tf_read[,i] == max(data.tf_read[,i])])
  }
  
  # Add to Seurat Object
  data.seurat$TF <- TF

  saveRDS(data.seurat, file = exp_out_seurat)
}

## Kneepoint filtering and TF assignment, specific for exp12-13, which contains combinations of 2 TFs
filtering_kneepoint_combination <- function(exp_cellranger, exp_tf_cell, exp_combination_list, exp_out_seurat){
  ## Cell Ranger
  data.cell_ranger <- Read10X_h5(filename = exp_cellranger, use.names = F)

  ## Cell Ranger barcodes
  data.barcodes <- unlist(data.cell_ranger@Dimnames[2])
  message(length(data.barcodes), " cell barcodes in cell ranger.")

  ## Create Seurat object
  data.seurat <- CreateSeuratObject(data.cell_ranger)
  rm(data.cell_ranger)

  # 10x
  data.tf_read <- fread(exp_tf_cell, data.table = F, )
  rownames(data.tf_read) <- data.tf_read$TFName
  data.barcodes_tf <- colnames(data.tf_read)[3:ncol(data.tf_read)]
  message(length(data.barcodes_tf), " cell barcodes in TF/Cell")
  
  # Overlap
  common.barcodes <- intersect(data.barcodes_tf, data.barcodes)
  data.tf_read <- data.tf_read[,common.barcodes] # Filtered TF Matrix
  message(length(common.barcodes), " cell barcodes are overlapping.")
  
  # Get combinations
  meta_combinations <- fread(exp_combination_list, data.table = F)
  rownames(meta_combinations) <- paste0(meta_combinations$TF1,"-",meta_combinations$TF2)
  
  # Precompute total TF read counts per cell barcode
  nCounts <- colSums(data.tf_read)
  
  # Calculate rate of main and second TF using Rfast package
  all.top2 <- colnth(as.matrix(data.tf_read), rep(2,ncol(data.tf_read)), descending = T, num.of.nths = 2)
  first.indexes <- colnth(as.matrix(data.tf_read), index.return = T, rep(1,ncol(data.tf_read)), descending = T, num.of.nths = 1) # Can't do both?
  second.indexes <- colnth(as.matrix(data.tf_read), index.return = T, rep(2,ncol(data.tf_read)), descending = T, num.of.nths = 1) # Can't do both?
  
  # Prepare output dataframe with rate of main and second TF
  data.tf_read.top2 <- data.frame(matrix(nrow = ncol(data.tf_read), ncol = 6), row.names = colnames(data.tf_read))
  colnames(data.tf_read.top2) <- c("TFmax", "TFmax.rate", "nMax", "TF2nd", "TF2nd.rate", "n2nd")
  
  # Fill it
  data.tf_read.top2$nMax <- all.top2[1,]
  data.tf_read.top2$TFmax.rate <- data.tf_read.top2$nMax / nCounts
  data.tf_read.top2$n2nd <- all.top2[2,]
  data.tf_read.top2$TF2nd.rate <- data.tf_read.top2$n2nd / nCounts
  data.tf_read.top2$TFmax <- rownames(data.tf_read)[first.indexes]
  data.tf_read.top2$TF2nd <- rownames(data.tf_read)[second.indexes]
  
  # Generate pseudo-TF as sum of top 2 TFs
  data.tf_read.top2$nPseudoTF <- data.tf_read.top2$nMax + data.tf_read.top2$n2nd
  data.tf_read.top2$PseudoTF.rate <- (data.tf_read.top2$nMax + data.tf_read.top2$n2nd) / nCounts
  
  # Remove top barcodes if n = 0
  data.tf_read.top2$TFmax[data.tf_read.top2$nMax == 0] <- NA
  data.tf_read.top2$TF2nd[data.tf_read.top2$n2nd == 0] <- NA
  
  # Which pair of TF could be a combination, based on metadata
  data.tf_read.top2$potential.combination <- paste0(data.tf_read.top2$TFmax,"-",data.tf_read.top2$TF2nd) %in% rownames(meta_combinations)
  
  # Filter cells with too low number of counts
  data.tf_read.top2 <- data.tf_read.top2[names(nCounts[nCounts > 5]),,drop=F]
  
  # Find the cutoff of the "Pseudo TF" thanks to kneepoint detection 
  PseudoTF.rate_order <- data.tf_read.top2[order(data.tf_read.top2$PseudoTF.rate, decreasing = T), "PseudoTF.rate", drop=F]
  cutoff <- kneepointDetection(PseudoTF.rate_order$PseudoTF.rate)
  
  # Filter cells based on cutoff kneepoint pseudo TF
  data.tf_read.top2 <- subset(data.tf_read.top2, PseudoTF.rate >= PseudoTF.rate_order[cutoff$MinIndex, "PseudoTF.rate"])
  
  # Now, find the cutoff of the "Main TF", thanks to kneepoint detection to identify singlets versus combinations (real and doublets)
  TFmax.rate_order <- data.tf_read.top2[order(data.tf_read.top2$TFmax.rate, decreasing = T), "TFmax.rate", drop=F]
  cutoff <- kneepointDetection(TFmax.rate_order$TFmax.rate)
  
  # Select cells based on cutoff kneepoint TF max rate of selected cells 
  data.tf_read.top2$singles <- data.tf_read.top2$TFmax.rate >= TFmax.rate_order[cutoff$MinIndex,"TFmax.rate"]
  
  # Remove Doublets (keep only combination and single TFs)
  data.tf_read.top2 <- subset(data.tf_read.top2, potential.combination == TRUE | singles == TRUE)
  
  # Assign TFs and combinations
  data.tf_read.top2$TF2nd[data.tf_read.top2$singles] <- NA
  data.tf_read.top2$TF <- meta_combinations[paste0(data.tf_read.top2$TFmax,"-",data.tf_read.top2$TF2nd), "combination"]
  data.tf_read.top2$TF[is.na(data.tf_read.top2$TF)] <- data.tf_read.top2$TFmax[is.na(data.tf_read.top2$TF)]
  
  # Subset cell_ranger matrix
  data.seurat <- data.seurat[,rownames(data.tf_read.top2)]
  
  # Assign TFs to Seurat Object
  data.seurat$TF <- data.tf_read.top2$TF
  sum(table(data.seurat$TF))
  
  message(ncol(data.seurat), " cells are passing the kneepoint filtering")
  
  saveRDS(data.seurat, file = exp_out_seurat)
}

```

## Kneepoint filtering

# Exp05 - Enriched

```{r}
filtering_kneepoint(
  exp_cellranger = "./cellranger/C3H10_10X_exp05_filtered_feature_bc_matrix.h5",
  exp_tf_cell = "./results/C3H10_10X_exp05_enriched.read_matrix.txt",
  exp_out_seurat = "./results/C3H10_10X_exp05_enriched.cellranger_kneepoint.rds"
)
```

# Exp06 - Enriched

```{r}
filtering_kneepoint(
  exp_cellranger = "./cellranger/C3H10_10X_exp06_filtered_feature_bc_matrix.h5",
  exp_tf_cell = "./results/C3H10_10X_exp06_enriched.read_matrix.txt",
  exp_out_seurat = "./results/C3H10_10X_exp06_enriched.cellranger_kneepoint.rds"
)
```

# Exp07 - Enriched

```{r}
filtering_kneepoint(
  exp_cellranger = "./cellranger/C3H10_10X_exp07_filtered_feature_bc_matrix.h5",
  exp_tf_cell = "./results/C3H10_10X_exp07_enriched.read_matrix.txt",
  exp_out_seurat = "./results/C3H10_10X_exp07_enriched.cellranger_kneepoint.rds"
)
```

# Exp08 - Enriched

```{r}
filtering_kneepoint(
  exp_cellranger = "./cellranger/C3H10_10X_exp08_filtered_feature_bc_matrix.h5",
  exp_tf_cell = "./results/C3H10_10X_exp08_enriched.read_matrix.txt",
  exp_out_seurat = "./results/C3H10_10X_exp08_enriched.cellranger_kneepoint.rds"
)
```

# Exp09 - Enriched

```{r}
filtering_kneepoint(
  exp_cellranger = "./cellranger/C3H10_10X_exp09_filtered_feature_bc_matrix.h5",
  exp_tf_cell = "./results/C3H10_10X_exp09_enriched.read_matrix.txt",
  exp_out_seurat = "./results/C3H10_10X_exp09_enriched.cellranger_kneepoint.rds"
)
```

# Exp10 - Enriched

```{r}
filtering_kneepoint(
  exp_cellranger = "./cellranger/C3H10_10X_exp10_filtered_feature_bc_matrix.h5",
  exp_tf_cell = "./results/C3H10_10X_exp10_enriched.read_matrix.txt",
  exp_out_seurat = "./results/C3H10_10X_exp10_enriched.cellranger_kneepoint.rds"
)
```

# Exp11 - Enriched

```{r}
filtering_kneepoint(
  exp_cellranger = "./cellranger/C3H10_10X_exp11_filtered_feature_bc_matrix.h5",
  exp_tf_cell = "./results/C3H10_10X_exp11_enriched.read_matrix.txt",
  exp_out_seurat = "./results/C3H10_10X_exp11_enriched.cellranger_kneepoint.rds"
)
```

# Exp12-13

There, it is special, since this study was sequenced twice (thus we need to merge the .h5 files). Moreover, the study contains combination TF (pairs of TFs), so we cannot run the kneepoint algorithm as is on these cells.

# First, let's create the merged matrix: C3H10_10X_exp12-13_enriched.read_matrix.txt

```{r}
# Reading exp12
TF_EXP12 <- fread("./results/C3H10_10X_exp12_enriched.read_matrix.txt", header = T, stringsAsFactors = F, data.table = F)
# Adding back the '-1', since some cell barcodes are the same between exp12 and exp13
colnames(TF_EXP12)[3:ncol(TF_EXP12)] <- paste0(colnames(TF_EXP12)[3:ncol(TF_EXP12)], "-1")
message(ncol(TF_EXP12) - 2, " cells in exp12 TF matrix")

# Reading exp13
TF_EXP13 <- fread("./results/C3H10_10X_exp13_enriched.read_matrix.txt", header = T, stringsAsFactors = F, data.table = F)
# Adding back the '-2', since some cell barcodes are the same between exp12 and exp13
colnames(TF_EXP13)[3:ncol(TF_EXP13)] <- paste0(colnames(TF_EXP13)[3:ncol(TF_EXP13)], "-2")
message(ncol(TF_EXP13) - 2, " cells in exp13 TF matrix")

# Merging
#all(TF_EXP12$TFName == TF_EXP13$TFName) # TRUE
#all(TF_EXP12$TFId == TF_EXP13$TFId) # TRUE
TF_EXP12_13 <- cbind(TF_EXP12, TF_EXP13[3:ncol(TF_EXP13)])

# Writing
fwrite(TF_EXP12_13, file = "./results/C3H10_10X_exp12-13_enriched.read_matrix.txt", sep = "\t", quote = F, row.names = F, col.names = T)

message(ncol(TF_EXP12_13) - 2, " cells in merged TF matrix")
rm(TF_EXP12, TF_EXP13, TF_EXP12_13)
```

## Kneepoint algo for exp12-13 - enriched

```{r}
filtering_kneepoint_combination(
  exp_cellranger = "./cellranger/C3H10_10X_exp12-13_filtered_feature_bc_matrix.h5",
  exp_tf_cell = "./results/C3H10_10X_exp12-13_enriched.read_matrix.txt",
  exp_combination_list = "./C3H10_10X_Metadata_exp12-13_combination.txt",
  exp_out_seurat = "./results/C3H10_10X_exp12-13_enriched.cellranger_kneepoint.rds")
```

