---
title: "1.6 Computing functional cells"
author: "Wangjie Liu / Vincent Gardeux"
date: "2024/01/04"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "./")
getwd()
```

## Libraries & functions

First, I'm loading the required libraries & functions

```{r}
suppressPackageStartupMessages(library(Seurat)) # For handling 10x h5 file
suppressPackageStartupMessages(library(data.table)) # For fread/fwrite functions
suppressPackageStartupMessages(library(doSNOW)) # For parallelization
suppressPackageStartupMessages(library(crayon)) # Just for bolding the console output :D

cat(bold("Seurat"), "version", as.character(packageVersion("Seurat")), "\n")
cat(bold("data.table"), "version", as.character(packageVersion("data.table")), "\n")
cat(bold("doSNOW"), "version", as.character(packageVersion("doSNOW")), "\n")

toReadableTime <- function(s){
  s <- round(s)
  if(s < 60) return(paste0(s, "s"))
  mn <- s %/% 60
  s <- s %% 60
  if(mn < 60) return(paste0(mn, "mn ", s, "s"))
  h <- mn %/% 60
  mn <- mn %% 60
  if(h < 24) return(paste0(h, "h ",  mn, "mn ", s, "s"))
  d <- h %/% 24
  h <- h %% 24
  return(paste0(d, "d ", h, "h ", mn, "mn ", s, "s"))
}

find_functional_cells <- function(seurat.object, threshold, threshold.non_con, threshold.con){
  # Normalizing and calculating PCA
  seurat.object <- NormalizeData(seurat.object, verbose = F) 
  seurat.object <- suppressWarnings(FindVariableFeatures(seurat.object, verbose = F, selection.method = "vst"))
  seurat.object <- ScaleData(seurat.object, verbose = F)
  n_pcs <- min(21, ncol(seurat.object))-1
  seurat.object <- RunPCA(seurat.object, npcs = n_pcs, features = VariableFeatures(seurat.object), verbose = F)
  pca <- seurat.object@reductions$pca@cell.embeddings
  
  # Centroids for all D0 in PCA space
  mean_D0s <- colMeans(pca[seurat.object$TF %in% c("D0", "D0_confluent"),])
  mean_D0s_non_confluent <- colMeans(pca[seurat.object$TF == "D0",])
  mean_D0s_confluent <- colMeans(pca[seurat.object$TF == "D0_confluent",])

  # Distance of all cells to centroids
  Dist_to_meanD0s <- apply(pca, 1, function(x){
      pos.vector <- rbind(x, mean_D0s)
      length.vectors <- dist(pos.vector, method = "euclidean", diag = F, upper = F, p = 2)[1] # p = 2, euclidean distance
      return(length.vectors)
  })
  
  Dist_to_meanD0s_non_confluent <- apply(pca, 1, function(x){
      pos.vector_non_confluent <- rbind(x, mean_D0s_non_confluent)
      length.vectors_non_confluent <- dist(pos.vector_non_confluent, method = "euclidean", diag = F, upper = F, p = 2)[1] # p = 2, euclidean distance
      return(length.vectors_non_confluent)
  })
  
  Dist_to_meanD0s_confluent <- apply(pca, 1, function(x){
      pos.vector_confluent <- rbind(x, mean_D0s_confluent)
      length.vectors_confluent <- dist(pos.vector_confluent, method = "euclidean", diag = F, upper = F, p = 2)[1] # p = 2, euclidean distance
      return(length.vectors_confluent)
  })
  
  # Compute threshold as quantile of D0 cells
  distance_thresholded_D0 <- quantile(Dist_to_meanD0s[seurat.object$TF %in% c("D0", "D0_confluent")], threshold)
  distance_thresholded_non_confluent <- quantile(Dist_to_meanD0s_non_confluent[seurat.object$TF %in% c("D0")], threshold.non_con)
  distance_thresholded_confluent <- quantile(Dist_to_meanD0s_confluent[seurat.object$TF %in% c("D0_confluent")], threshold.con)
  
  # Extract FUNCTIONAL cells, i.e. cells that pass the threshold
  cells.oi <- names(Dist_to_meanD0s)[Dist_to_meanD0s > distance_thresholded_D0]
  cells.oi_non_confluent <- names(Dist_to_meanD0s_non_confluent)[Dist_to_meanD0s_non_confluent > distance_thresholded_non_confluent]
  cells.oi_confluent <- names(Dist_to_meanD0s_confluent)[Dist_to_meanD0s_confluent > distance_thresholded_confluent]
  cells.funct <- intersect(intersect(cells.oi, cells.oi_non_confluent), cells.oi_confluent)
    
  # Non D0 FUNCTIONAL cells
  cells.funct <- names(seurat.object$TF[cells.funct] )[!seurat.object$TF[cells.funct] %in% c("D0", "D0_confluent")]
  return(cells.funct)
}
```

# Adding a few metadata to prepare the calculation
```{r}
# Reading integrated Seurat object from 1.5.3
data.seurat <- readRDS("results/C3H10_10X_all_exps_merged_genefiltered_integrated.rds")

# Creating log_vector and putting D0 vector to 0
data.seurat$log_vector <- data.seurat$Log_Vector_UMI
data.seurat$log_vector[data.seurat$TF == "D0" | data.seurat$TF == "D0_confluent"] <- 0  # D0 178, D0_confluent 166, no mCherry_BC14

# Creating the batch_overall, combining exp that were processed in the same batch
data.seurat$batch_overall <- "batch1" # exp05 and exp06
data.seurat$batch_overall[data.seurat$batch %in% c("exp07","exp08")] <- "batch2" 
data.seurat$batch_overall[data.seurat$batch %in% c("exp09")] <- "batch3" 
data.seurat$batch_overall[data.seurat$batch %in% c("exp10","exp11")] <- "batch4" 
data.seurat$batch_overall[data.seurat$batch %in% c("exp12-13")] <- "batch5" 
table(data.seurat$batch_overall)

# Fixing metadata
data.seurat$orig.ident <- data.seurat$batch
```

# Running the main parallel code for computing functional cells
**Goal:** Identify bloby versus functional cells per TF
**Method:** Search euclidean distance of TF cells to D0 +/ D0confluent in PCA space to identify functional versus non functional cells
```{r}
# Parameters
percentile.D0s <- 0.8
percentile.D0noncon <- 0.8
percentile.D0con <- 0.8
num_cores <- 32

# I will proceed each phase, each batch, each TF separately
allPhases <- c("G1","S","G2M")
allBatches <- c("batch1","batch2","batch3","batch4","batch5")
allTFs <- as.character(unique(data.seurat$TF))
allTFs <- allTFs[!allTFs %in% c("D0", "D0_confluent", "MatureAdipo")]
message("There are ", length(allTFs), " TFs to process")

# All possible combinations
list_params <- list()
l <- 0
for(p in allPhases){
  for(b in allBatches){
    for(t in allTFs){
      l <- l + 1
      list_params[[l]] <- list(phase = p, batch = b, tf = t)
    }
  }
}
message("There are ", length(list_params), " combinations to run")

# Set the number of cores/workers to use
cl <- makeCluster(num_cores)
registerDoSNOW(cl)

progress.bar <- txtProgressBar(style = 3, min = 1, max = length(list_params))

# Perform computation for each combination of parameters (in parallel)
start_time = Sys.time()

funct_annotations <- foreach(i = 1:length(list_params), .packages = c("Seurat"), .combine = "c", .options.snow = list(progress = function(x) setTxtProgressBar(progress.bar, x))) %dopar% {
  condition <- data.seurat$Phase == list_params[[i]]$phase & data.seurat$batch_overall == list_params[[i]]$batch & data.seurat$TF %in% c("D0", "D0_confluent", list_params[[i]]$tf)
  if(sum(condition) == 0) return(NULL)
  return(find_functional_cells(seurat.object = data.seurat[,condition], threshold = percentile.D0s, threshold.non_con = percentile.D0noncon, threshold.con = percentile.D0con))
}

close(progress.bar)
stopCluster(cl)

message("Total time = ", toReadableTime(as.numeric(Sys.time() - start_time, unit = "secs")))

message(length(funct_annotations), " functional cells were found") # 19389 in final atlas # 19541 from my data
```

# Updating the Seurat object
```{r}
data.seurat$functional_cells <- F
data.seurat$functional_cells[funct_annotations] <- T
data.seurat$functional_cells[data.seurat$TF == "MatureAdipo"] <- "MatureAdipo"
data.seurat$functional_cells[data.seurat$TF %in% c("D0", "D0_confluent")] <- "D0"

saveRDS(data.seurat, "results/C3H10_10X_all_exps_merged_genefiltered_integrated_functional.rds")
```
