---
title: "Figure S4a-c - Reproducibility script"
author: "Wangjie Liu"
date: "2024-03-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "./")
getwd()
```

## Libraries & functions

First, I'm loading the required libraries & functions

```{r}
suppressPackageStartupMessages(library(Seurat)) # For single-cell pipeline
suppressPackageStartupMessages(library(ggplot2)) # For plotting
suppressPackageStartupMessages(library(tidyverse)) # For data tidying
suppressPackageStartupMessages(library(clustree)) # For clustering tree
suppressPackageStartupMessages(library(crayon)) # Just for bolding the console output :D

cat(bold("Seurat"), "version", as.character(packageVersion("Seurat")), "\n")
cat(bold("ggplot2"), "version", as.character(packageVersion("ggplot2")), "\n")
cat(bold("tidyverse"), "version", as.character(packageVersion("tidyverse")), "\n")
cat(bold("clustree"), "version", as.character(packageVersion("clustree")), "\n")
```

## read functional TF atlas with functional TF cells, D0 and ref cells in G1 at default phase

```{r}
seu <- readRDS("results/C3H10_10X_all_exps_integrated_functional_TF_atlas_G1_Phase.rds")
```

# Figure S4a

```{r}
# find clusters for the functional TF atlas
# run different resolution and then clustree to find an optimal resolution for clustering
for (i in seq(0.1, 1.5, 0.1)){
    seu <- FindClusters(seu, resolution = i, verbose = FALSE)
    p <- DimPlot(seu, group.by = "seurat_clusters",label = T)+labs(title = paste0("resolution_",i))
    print(p)
}
```

```{r Fig_S4a, fig.width=10, fig.height=14}
p.clustree <- clustree(seu, node_text_size = 0, edge_width = 1)
p.clustree
ggsave(p.clustree, filename = "figures/Fig_S4a.pdf",
       width = 10, height = 14, units = "in")
```

# Figure S4b

```{r Fig_S4b, fig.width=7, fig.height=5}
# choose resolution 1.0 for clustering, because the number of clusters becomes relatively fixed at res 0.9-1.0-1.1 and the intercluster arrows/interactions are relatively less when changing res from 1.0 to 1.1 
seu$seurat_clusters <- seu$integrated_snn_res.0.6

# set color scheme
Cols_res_0.6 <- c("#FFFFCC","#FFCCCC","#99FF99","#FFCC99","#FF9933","#339900",
                "#CCCC00","#FFFF00","#CC6666","#FF99FF","#CC00FF","#660066",
                "#66CCCC","#003399","#CC99CC","#3399FF","#CC9966","#99CCFF","#FF9999")

names(Cols_res_0.6) <- as.character(0:18)

# plot the atlas on UMAP
p_umap <- DimPlot(seu, group.by = "seurat_clusters", order = T, label = F, cols = Cols_res_0.6)+
    theme(axis.line = element_blank(),
        legend.position = "none", 
        panel.grid.major=element_blank(), 
        panel.grid.minor=element_blank(), 
        axis.text=element_blank(), 
        axis.ticks=element_blank(), 
        plot.background=element_blank(), 
        panel.background=element_blank())+
  xlab(label = "") + ylab(label = "")+
  labs(title = NULL)
p_umap
ggsave(p_umap, filename = "figures/Fig_S4b.pdf", width = 7, height = 5)
```

# Figure S4c

### extract clustering metadata at "integrated_snn_res.0.6"

```{r}
df <- data.frame(row.names = rownames(seu@meta.data),
                 TF = seu$TF,
                 Cluster = seu$integrated_snn_res.0.5)
df$Cluster <- as.factor(df$Cluster)
```

### filter out TFs with less than 30 cells

```{r}
df <- df %>% group_by(TF) %>% mutate(Number_of_TF_cells = length(TF))
df <- df[df$Number_of_TF_cells >= 30,] 
```

### calculate the number of functional cells per TFs per clusters, filter out the one with less than 10 cells

```{r}
df <- lapply(unique(df$Cluster), function(x){
  df <- df[df$Cluster == x, ] %>% group_by(TF) %>% mutate(Number_of_TF_cells_in_each_clusters = length(TF))
  df <- df[df$Number_of_TF_cells_in_each_clusters >= 10,] # minimum TF cells in each cluster
  return(df)
})
df <- data.table::rbindlist(df)
```

### calculate the percentage of cells per TF per cluster, filter out the one with less than 15% of functional TF cells or greater than 85% of functional TF cells

```{r}
df_percent <- lapply(unique(df$Cluster), function(x){
  Percentage_TF <- df[df$Cluster == x, ] %>% group_by(TF) %>% mutate(percentage_TF = length(TF)/Number_of_TF_cells)
  Percentage_TF <- Percentage_TF[!duplicated(Percentage_TF),]
  return(Percentage_TF)
})
df_percent <- data.table::rbindlist(df_percent)
df_percent$percentage_TF <- df_percent$percentage_TF*100 %>% as.numeric()
df_percent$percentage_TF <- round(df_percent$percentage_TF)
df_percent <- df_percent[df_percent$percentage_TF >= 15 & df_percent$percentage_TF < 85,]
```

### filter out TFs involved in only 1 cluster of G1 and non-control clusters 

```{r}
D0.clusters <- df_percent$Cluster[df_percent$TF == "D0" & df_percent$percentage_TF >= 15] %>% as.character() # main clusters of D0 (non-confluent only)
D0.clusters # "5" "1" "3"
D0.confluent.clusters <- df_percent$Cluster[df_percent$TF == "D0_confluent" & df_percent$percentage_TF >= 15] %>% as.character()
D0.confluent.clusters # "1" "0" "3"
## use D0_confluent as the real control
TF.keep <- table(df_percent$TF[!df_percent$Cluster %in% D0.confluent.clusters]) 
TF.keep <- TF.keep[TF.keep > 1] %>% names()
df_percent <- df_percent[df_percent$TF %in% c(TF.keep, "D0","D0_confluent"),]
```

### prepare for heatmap plot

```{r}
nCluster <- unique(df_percent$Cluster) %>% as.character()
nCluster %>% length() # 17
nCluster <- nCluster[order(as.numeric(nCluster), decreasing = F)]
# reform the data frame for heatmap plot
df_mat <- lapply(unique(df_percent$TF), function(x){
  df.reform <- data.frame(row.names = df_percent$Cluster[df_percent$TF == x],
                          percentage_TF = df_percent$percentage_TF[df_percent$TF == x])
  Cluster.supplement <- nCluster[!nCluster %in% rownames(df.reform)]
  df.reform <- rbind(df.reform, data.frame(row.names = Cluster.supplement,
                                           percentage_TF = rep(0, length(Cluster.supplement)))) 
  df.reform <- t(df.reform) %>% as.data.frame()
  df.reform$TF <- x
  df.reform <- df.reform[, c(as.character(nCluster),"TF")]
  return(df.reform)
})


df_mat <- data.table::rbindlist(df_mat) # as.matrix here convert percentage to characters
rownames.df_mat <- df_mat$TF
df_mat$TF <- NULL
dim(df_mat) # 19 17
df_mat <- as.matrix(df_mat)
rownames(df_mat) <- rownames.df_mat

#remove TF combination, adipo and myo ref cells
TF.com <- c("Cebpa-Mycn","Cebpa-Myog","Cebpa-Pparg","Mycn-Myog","Mycn-Pparg","Mycn-Runx2","Pparg-Runx2")
df_mat <- df_mat[!rownames(df_mat) %in% c(TF.com, "Adipo_ref","Myo_ref"),]
```

### plot heatmap

```{r Fig_S4b, fig.width=5, fig.height=6}
#pheatmap
p_heatmap <- pheatmap::pheatmap(df_mat, 
                                cellheight = 10,
                                cellwidth = 12, 
                                fontsize_row = 10, 
                                fontsize_col = 10,
                                color = colorRampPalette(c("#edf8e9","#006d2c"))(100),
                                border_color = NA,
                                cluster_rows = T, 
                                clustering_method = "ward.D2",
                                cluster_cols = F)

p_heatmap

ggsave(p_heatmap, filename = "figures/Fig_S4c.pdf",
       width = 5, height = 6, units = "in")
```


